---
title: "Source Attribution Analysis for C. jejuni from Southern Alberta, Canada"
output: github_document
author: Benjamin Hetman
date: "`r format(Sys.Date())`"
---

# Source attribution analysis for *C. jejuni* from southern Alberta, Canada

## Step 0: Load libraries and custom functions

For this step, it is often useful to write a separate helper file. It keeps the present workspace organized by reducing clutter. Here we'll use the `source()` command to call another file labelled `helper.R` which contains many of our custom functions.

```{r source helper script, message=FALSE, warning=FALSE}
source("helper.R")

```

## Step 1: Load data from MS Excel file

```{r load data}

campy_raw <- readxl::read_xlsx(here("DATA","SWA_data_for_attribution.xlsx"), sheet = 1, col_names = TRUE, guess_max = 1000, na=c("?", "#N/A"))

```

------------------------------------------------------------------------

## Step 2: Verify data

```{r explore and verify data}

campy_human <- campy_raw %>%  filter(Source == "Human")
campy_nonhuman <- campy_raw %>%  filter(Source != "Human")

```

Summarize the human data set by year, providing some descriptive statistics.

```{r summarize human data}

# Summarize Human case data by year
campy_human %>%
  group_by(Year) %>%  
  summarize(n.isolates = n(), 
            n.subtypes = length(unique(BCM95.v2)), 
            min.size = min(BCM95.SA.size, na.rm = TRUE), 
            max.size = max(BCM95.SA.size, na.rm = TRUE),
            subtype.size = list(BCM95.SA.size[!is.na(BCM95.SA.size)]),
            .groups = "keep") %>%
  mutate(Year = as.factor(Year)) %>% 
  flextable() %>%  
  compose(j="subtype.size", value = as_paragraph(
    plot_chunk(value = subtype.size, type = "dens", col = "blue")
  )) %>%
  set_caption("Summary of human C.jejuni isolates") 
```

Summarize the non-human data, also by year.

```{r summarize non-human data}

# Summarize Human case data by year
campy_nonhuman %>%
  group_by(Year, Source) %>%  
  summarize(n.isolates = n(), 
            n.subtypes = length(unique(BCM95.v2)), 
            min.size = min(BCM95.SA.size, na.rm = TRUE), 
            max.size = max(BCM95.SA.size, na.rm = TRUE),
            subtype.size = list(BCM95.SA.size[!is.na(BCM95.SA.size)]),
            .groups = "keep") %>%  
  mutate(Year = as.factor(Year)) %>%  
  flextable() %>%  
  compose(j="subtype.size", value = as_paragraph(
    plot_chunk(value = subtype.size, type = "dens", col = "red")
  )) %>%
  set_caption("Summary of non-human C.jejuni isolates")  
  
  
```

## Step 3: Set up randomization for Source Attribution

```{r Monte Carlo Source Attribution analysis, fig.width=8}

# Notes: 
# Purpose of this analysis is to generate estimate and 95% confidence intervals
# surrounding the attribution of source for each unique subtype. 

# The objective will be to iterate through each unique subtype, and randomly sample 
# a subset (i.e., subsample) to calculate the estimated primary source frequency, 
# then calculate a confidence interval around that distribution. 


f <- function(human.data, nonhuman.data, iterations) {
  # Initialize list to store results
  results_table <- tibble(
    subtype = unique(human.data$BCM95.v2),
    n.human = NA,
    n.nonhuman = NA,
    Chicken = NA,
    Chicken.low = NA,
    Chicken.high = NA,
    Cattle = NA,
    Cattle.low = NA,
    Cattle.high =  NA,
    Water =  NA,
    Water.low = NA,
    Water.high = NA,
    )
  
  # Loop through subtypes from the human dataset
  for(j in unique(human.data$BCM95.v2)){
    #Subset non-human data based on human subtype
    p <- nonhuman.data %>% filter(BCM95.v2 == j)
    q <- human.data %>%  filter(BCM95.v2 == j) 
    
    results_table[results_table$subtype == j,]$n.nonhuman <- nrow(p)
    results_table[results_table$subtype == j,]$n.human <- nrow(q)
    #If less than 5 isolates in non-human subset, don't calculate stats and move
    # to the next subtype. 
    if(nrow(p) < 5) {
      next
    }
    # Use 50% of isolates from subtype to sample
    n <- round(nrow(p) * 0.5)
    
    # Initialize empty tibble and draw random samples
    x <- tibble(n.Cattle = numeric(), n.Chicken =numeric(), n.Water=numeric())
    for(i in 1:iterations){
      sub <- sample_n(p, n, replace = FALSE)
      x[i, "n.Chicken"] <- sub %>%  filter(Source == "Chicken") %>%  nrow()
      x[i, "n.Cattle"] <- sub %>%  filter(Source == "Cattle") %>%  nrow()
      x[i, "n.Water"] <- sub %>%  filter(Source == "Water") %>%  nrow()
    }
      
    # Calculate descriptive statistics
    #----------------------------------------------------------------------------
      mean.chicken <- mean(x$n.Chicken)
      mean.cattle <- mean(x$n.Cattle)
      mean.water <- mean(x$n.Water)
      sd.chicken <- sd(x$n.Chicken)
      sd.cattle <- sd(x$n.Cattle)
      sd.water <- sd(x$n.Water)
     #----------------------------------------------------------------------------
      ci.hi.chicken <- mean.chicken + 1.96*(sd.chicken/sqrt(length(x$n.Chicken)))
      ci.low.chicken <- mean.chicken - 1.96*(sd.chicken/sqrt(length(x$n.Chicken)))
      
      ci.hi.cattle <- mean.cattle + 1.96*(sd.cattle/sqrt(length(x$n.Cattle)))
      ci.low.cattle <- mean.cattle - 1.96*(sd.cattle/sqrt(length(x$n.Cattle)))
      
      ci.hi.water <- mean.water + 1.96*(sd.water/sqrt(length(x$n.Water)))
      ci.low.water <- mean.water - 1.96*(sd.water/sqrt(length(x$n.Water)))
     #----------------------------------------------------------------------------
      # Store results in final table:
      results_table[results_table$subtype == j,]$Chicken = round(mean.chicken/n, 2)
      results_table[results_table$subtype == j,]$Chicken.high = round(ci.hi.chicken/n, 2)
      results_table[results_table$subtype == j,]$Chicken.low = round(ci.low.chicken/n, 2)
      results_table[results_table$subtype == j,]$Cattle = round(mean.cattle/n, 2)
      results_table[results_table$subtype == j,]$Cattle.high =  round(ci.hi.cattle/n, 2)
      results_table[results_table$subtype == j,]$Cattle.low = round(ci.low.cattle/n, 2)
      results_table[results_table$subtype == j,]$Water =  round(mean.water/n, 2)
      results_table[results_table$subtype == j,]$Water.high = round(ci.hi.water/n, 2)
      results_table[results_table$subtype == j,]$Water.low = round(ci.low.water/n, 2)
  }
return(results_table)
}


subtype.table <- f(campy_human, campy_nonhuman, 100) 
subtype.table <- arrange(subtype.table, desc(n.human))



flextable(subtype.table)


```

```{r}

# Set up random sampler for human data only: 
human_sampler <- function(human.data, subtype.data, iterations, sample.n){
  
  subtype.data <- subtype.data %>%  filter(!is.na(Chicken))
  human.data <- human.data %>%  filter(BCM95.v2 %in% subtype.data$subtype)
  
  results_list <- list()
  for(i in 1:iterations){
  x <- 
    sample_n(human.data, sample.n) %>%  
    select(Key, BCM95.v2) %>%
    group_by(BCM95.v2) %>%  
    summarize(n.human = n()) %>%  
    left_join(subtype.data, by=c("BCM95.v2"="subtype")) %>% 
    select(-n.human.y, -n.nonhuman) %>% 
    mutate(across(Chicken:Water.high, ~ .x * n.human.x)) %>% 
    summarize(across(n.human.x:Water.high, ~ sum(.x, na.rm=TRUE)))
  
  results_list[[i]] <- x  
  }
  return(bind_rows(results_list))
}

sa.results <- human_sampler(campy_human, subtype.table, 30, 500)
flextable(sa.results)

```


```{r}

start.df <- 
 sa.results %>% 
  select(-n.human.x) %>% 
  rowid_to_column("Index")

chicken.df <- start.df %>% 
  select(Index, starts_with("Chicken")) %>%  
  mutate(Source = "Chicken") %>% 
  rename(c("Index"="Index", "Mean"="Chicken", "Low"="Chicken.low", "High"="Chicken.high", "Source"="Source"))

cattle.df <- start.df %>% 
  select(Index, starts_with("Cattle")) %>%  
  mutate(Source = "Cattle") %>% 
  rename(c("Index"="Index", "Mean"="Cattle", "Low"="Cattle.low", "High"="Cattle.high", "Source"="Source"))

water.df <- start.df %>% 
  select(Index, starts_with("Water")) %>%  
  mutate(Source = "Water") %>% 
  rename(c("Index"="Index", "Mean"="Water", "Low"="Water.low", "High"="Water.high", "Source"="Source"))
  
plot.df <- bind_rows(chicken.df, cattle.df, water.df) %>%  
  arrange(Index)


ggplot(plot.df, aes(x = Index, y = Mean)) + 
  geom_bar(stat = "identity", aes(fill = Source)) +
  theme_minimal() + 
  labs(title = "Distribution of attributed source of n=500 human clinical isolates randomly sampled 100 times from southern Alberta, Canada") +
  scale_fill_manual(values = c("#a8a355","#d4674c", "#67727e"))


```

















